Class differences in social networks: Evidence from a referral experiment

### 2025-06-16 writing/manuscript.tex
v0 - results ctd.

\title{Class differences in social networks: Evidence from a referral experiment}
% \\ \large Experimental Design \\
% \title{Skills, Inequality, and Referrals}
\author{Manuel Munoz,\footremember{liser}{Luxembourg Institute of Socio-Economic Research} Ernesto Reuben,\footnote{Division of Social Science, New York University Abu Dhabi} \footrecall{liser} Reha Tuncer\footnote{University of Luxembourg}}  
% \footnote{Center for Behavioral Institutional Design at New York University Abu Dhabi} 
\begin{document}
\linenumbers 
% HERE FOR LINES
\maketitle

@@

So far we have looked at the entire network without considering the intensity of connections between students. In our network data set, this variable amounts to the number of classes taken together with peers. As we will see in the next section, referrals go to peers with whom participants have taken a lot of courses with \textcolor{red}{on average XX}, implying the intensity of the connection matters. We begin by dissecting what the intensity means in our context. As students take more courses together, the proportion of peers from the same academic program quickly goes beyond 95\% (see Figure \ref{share_same_program_by_tie}). Similarly, the average network size drops very quickly from above 210 to around 25 (see Figure \ref{conns_by_tie}). Both results indicate that actual referral considerations orginate from a much smaller pool of individuals from the same academic program.

@@

We look at how SES groups are distributed across programs to see evidence of SES-based selection (see Figure \ref{ses_distribution_by_fees}). Indeed, low-SES select into more affordable programs, followed by middle-SES. High-SES sort almost exclusively in above average costing programs, with a third of students selecting into medicine and creating a very skewed distribution. With this finding, program selection could be the reason why low- and high-SES networks tend to segregate as the number of courses taken increases. The next section characterizes the referrals, and we will return to the diversity implications of program selection once we understand how referrals were made.

@@

We observe 1342 referrals in our final data set. More than 90\% of these consist from participants referring for both areas of the national entry exam, as can be seen in Table \ref{tab:referral_distribution}.

What are the characteristics of the individuals who receive referrals, and how do they compare to others in the enrollment network?

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Summary statistics for network members by nomination status}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccccc}
  \toprule
  & \multicolumn{2}{c}{Verbal} & \multicolumn{2}{c}{Math} \\
  \cmidrule{2-3} \cmidrule{4-5}
  & Not Referred & Referred & Not Referred & Referred \\
  \midrule
  Reading z-score & 0.070 & 0.509 & 0.079 & 0.465 \\
  & (0.003) & (0.039) & (0.003) & (0.040) \\
  [1ex]
  Math z-score & 0.079 & 0.452 & 0.087 & 0.590 \\
  & (0.003) & (0.042) & (0.003) & (0.043) \\
  [1ex]
  GPA z-score & -0.066 & 0.705 & -0.069 & 0.711 \\
  & (0.003) & (0.041) & (0.003) & (0.041) \\
  [1ex]
  Tie strength z-score & -0.153 & 2.690 & -0.184 & 2.488 \\
  & (0.003) & (0.091) & (0.003) & (0.090) \\
  [1ex]
  Low-SES & 0.334 & 0.374 & 0.338 & 0.384 \\
  & (0.001) & (0.019) & (0.001) & (0.019) \\
  [1ex]
  Med-SES & 0.515 & 0.513 & 0.513 & 0.507 \\
  & (0.001) & (0.019) & (0.001) & (0.019) \\
  [1ex]
  High-SES & 0.151 & 0.113 & 0.149 & 0.109 \\
  & (0.001) & (0.012) & (0.001) & (0.012) \\
  % [1ex]
  % Female & 0.524 & 0.562 & 0.520 & 0.428 \\
  % & (0.001) & (0.019) & (0.001) & (0.019) \\
  % [1ex]
  % Age & 20.945 & 20.501 & 20.961 & 20.490 \\
  % & (0.006) & (0.104) & (0.006) & (0.099) \\
  \midrule
  Observations & 128,174 & 673 & 127,481 & 669 \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} Standard errors in parentheses. GPA, test scores, and tie strength are standardized at the network level. For each referrer's network, we first calculated the mean and standard deviation of each measure. We then computed the average of these means and standard deviations across all referrers. Each individual's score was standardized using these network-level statistics. The standardization formula is $z = (x - \bar{x}_{network}) / \sigma_{network}$, where $\bar{x}_{network}$ and $\sigma_{network}$ are the average of network means and standard deviations, respectively. Low-SES, Med-SES, and High-SES are binary variables indicating the share of participants in estrato 1 and 2, 3 and 4, or 5 and 6, respectively. Tie strength measures the number of connections between individuals.
  \end{tablenotes}
  \label{tab:stats_by_nomination}
  \end{threeparttable}
\end{table}

@@

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Selection into the experiment}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc c@{}}
  \toprule
  & \multicolumn{1}{c}{\textbf{University}} & \multicolumn{1}{c}{\textbf{Sample}} & \multicolumn{1}{c}{\textit{\textbf{p}}} \\
  \midrule
  Reading score & 62.651 & 65.183 & 0.000 \\
  Math score & 63.973 & 67.477 & 0.000 \\
  GPA & 3.958 & 4.012 & 0.000 \\
  Low-SES & 0.343 & 0.410 & 0.000 \\
  Med-SES & 0.505 & 0.499 & 0.763 \\
  High-SES & 0.153 & 0.091 & 0.000 \\
  Female & 0.567 & 0.530 & 0.060 \\
  Age & 21.154 & 20.651 & 0.000 \\
  \midrule
  Observations & 4,417 & 734 & 5,151 \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} This table compares characteristics between the university and the experimental sample. $p$-values for binary outcomes (Low-SES, Med-SES, High-SES, Female) are from two-sample tests of proportions; for continuous variables, from two-sample $t$-tests with unequal variances. All reported \textit{p}-values are two-tailed.
  \end{tablenotes}
  \label{tab:selection}
  \end{threeparttable}
\end{table}

>>>>

\title{Class differences in social networks: Evidence from a referral experiment}
% \\ \large Experimental Design \\
% \title{Skills, Inequality, and Referrals}
\author{Manuel Munoz\footremember{liser}{Luxembourg Institute of Socio-Economic Research}, Ernesto Reuben\footnote{Division of Social Science, New York University Abu Dhabi} \footrecall{liser}, Reha Tuncer\footnote{University of Luxembourg}}  
% \footnote{Center for Behavioral Institutional Design at New York University Abu Dhabi} 
\begin{document}
\linenumbers 
% HERE FOR LINES
\maketitle

@@

So far we have looked at the entire network without considering the intensity of connections between students. In our network data set, this variable amounts to the number of classes taken together with peers. As we will see in the next section, referrals go to peers with whom participants have taken on average 14 courses courses with, implying the intensity of the connection matters. We begin by dissecting what the intensity means in our context. As students take more courses together, the proportion of peers from the same academic program quickly goes beyond 95\% (see Figure \ref{share_same_program_by_tie}). Similarly, the average network size drops very quickly from above 210 to below 50 (see Figure \ref{conns_by_tie}). Both results indicate that actual referral considerations orginate from a much smaller pool of individuals from the same academic program.

@@

We look at how SES groups are distributed across programs to see evidence of SES-based selection (see Figure \ref{ses_distribution_by_fees}). Indeed, low-SES select into more affordable programs, followed by middle-SES. High-SES sort almost exclusively in above average costing programs, with a third of students selecting into medicine and creating a very skewed distribution \textcolor{red}{KS test here}. With this finding, program selection could be the reason why low- and high-SES networks tend to segregate as the number of courses taken increases. The next section characterizes the referrals, and we will return to the diversity implications of program selection once we propose an understanding of how referrals were made.

@@

We observe 1342 referrals from our 734 participants in our final data set. More than 90\% of these consist from participants referring for both areas of the national entry exam (see Appendix Table \ref{tab:referral_distribution}). While each participant made one referral for Math and Reading parts of the exam, we pool these referrals in our main analysis and look at averages as there is no difference in terms of referral SES compositions between exam areas (see Appendix Table \ref{tab:referral_area}).

What are the characteristics of the individuals who receive referrals, and how do they compare to others in the enrollment network? Because we have an entire pool of potential candidates with one referral chosen from it, we compare the distributions for our variables of interest between the referred and non-referred students. First, referrals go to peers with whom the referrer has taken around 14 courses with on average, compared to almost 4 on average with others in their network (see Figure \ref{tie_hist}). This difference of 10.1 courses is significant ($t = 34.98$, $p < 0.001$), indicating that referrers choose individuals with whom they have stronger ties. While the median referral recipient has taken 12 courses together with the referrer, the median network member has shared only 2.8 courses. The interquartile range for referrals spans from 7.5 to 19.5 courses, compared to just 2.1 to 4.0 courses for the broader network, highlighting the concentration of referrals among peers within same program (93\%).

\begin{figure}[H]
  \centering
  \caption{Courses taken together with network members and referrals}\label{tie_hist}
  \includegraphics[width=0.6\linewidth]{../figures/tie_hist.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure compares the distributions of the number of courses taken together between referrers and their network members (orange) versus referrers and their chosen referral recipients (dark blue) for all 734 participants. 75\% of referral recipients having taken more than 7.5 courses together with the referrer, compared to only 25\% of network members. The distributions are significantly different (Kolmogorov-Smirnov test $D = 33.37$, $p < 0.001$).
  \end{tablenotes}
\end{figure}

Second, we examine entry exam score differences between referred students and the broader network. Referrals go to peers with an average score of 69.5 points, compared to 64.5 points for other network members (see Figure \ref{other_score_hist}). This difference of 5 points is significant ($t = 18.97$, $p < 0.001$), indicating that referrers choose higher-performing peers. While the median referral recipient scores 71.0 points, the median network member scores 65.1 points. The interquartile range for referrals spans from 65.5 to 75.0 points, compared to 63.5 to 66.9 points for the broader network, highlighting the clear concentration of referrals among higher performing peers.

\begin{figure}[H]
  \centering
  \caption{Entry exam scores of network members and referrals}\label{other_score_hist}
  \includegraphics[width=0.6\linewidth]{../figures/other_score_hist.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure compares the distributions of entry exam scores (Math and Reading average) between referrers' network members (orange) versus their chosen referral recipients (dark blue) for all 734 participants. 75\% of referral recipients score above 65.5 points compared to only 25\% of network members scoring above 66.9 points. The distributions are significantly different (Kolmogorov-Smirnov test $D = 71.16$, $p < 0.001$). 
  \end{tablenotes}
\end{figure}

Third, we examine the overall SES-compositions in referral selection. Referrals to low-SES peers constitute 37.9\% of all referrals, compared to 33.7\% low-SES representation in individual networks (see Table \ref{tab:ses_composition}). This over-representation of 4.3 pp. is marginally significant ($z = -1.70$, $p = 0.088$), suggesting a small overall tendency to favor low-SES peers. For middle-SES students, referrals constitute 51.0\% versus 51.4\% network representation (difference: -0.5 pp., $z = 0.17$, $p = 0.863$). High-SES referrals account for 11.1\% compared to 14.9\% network share, a significant under-representation of 3.8 pp. ($z = 2.17$, $p = 0.030$). These findings indicate that referral selection in general modestly deviates from proportional representation, with referrers showing a slight preference for lower-SES at the expense of high-SES peers relative to their network shares.

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{SES composition: network shares versus referral rates}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc c@{}}
  \toprule
   & \multicolumn{1}{c}{\textbf{Network average}} & \multicolumn{1}{c}{\textbf{Referrals}} & \multicolumn{1}{c}{\textit{\textbf{p}}} \\ 
  \midrule
  Low-SES & 33.7\% & 37.9\% & 0.088 \\
  Mid-SES & 51.4\% & 51.0\% & 0.863 \\
  High-SES & 14.9\% & 11.1\% & 0.030 \\
  \midrule
  N & 734 & 734 & \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} This table compares the average SES compositions of referrer enrollment networks and actual referral choices. $p$-values are from two-sample proportion tests. Network percentages represent the average SES group shares across individual networks. Referral percentages represent the actual SES distribution of chosen referrals.
  \end{tablenotes}
  \label{tab:ses_composition}
  \end{threeparttable}
\end{table}

\textcolor{red}{to do: 1- SES composition by SES, 2- expost choice sets 3- effect of the bonus 4- bias regression }

Table \ref{tab:stats_by_nomination}

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Summary statistics for network members by referral status}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccc}
  \toprule
  & \multicolumn{1}{c}{\textbf{\textcolor{red}{Not Referred - to remove}}} & \multicolumn{1}{c}{\textbf{Baseline Referred}} & \multicolumn{1}{c}{\textbf{Bonus Referred}} & \multicolumn{1}{c}{\textbf{\textit{p}}} \\
  \midrule
  Reading score & 64.212 & 68.044 & 67.180 & 0.063 \\
  Math score & 66.361 & 71.119 & 70.240 & 0.143 \\
  GPA & 3.909 & 4.158 & 4.148 & 0.583 \\
  Tie strength & 3.162 & 14.062 & 14.236 & 0.732 \\
  Low-SES & 0.336 & 0.366 & 0.393 & 0.304 \\
  Med-SES & 0.514 & 0.548 & 0.468 & 0.004 \\
  High-SES & 0.150 & 0.086 & 0.138 & 0.002 \\
  \midrule
  Observations & 255,655 & 699 & 643 \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} P-values are from two-sample t-tests with unequal variances for continuous variables and two-sample tests of proportions for binary variables, comparing Baseline Referred vs Bonus Referred groups. Reading and math scores are raw test scores. GPA is the grade point average. Tie strength measures the weighted frequency of contact between individuals. Low-SES, Med-SES, and High-SES are binary variables indicating the share of participants in estrato 1 and 2, 3 and 4, or 5 and 6, respectively. Not Referred includes all network members who were not nominated for referral. Baseline Referred and Bonus Referred include only those network members who were actually nominated for referral in each treatment condition.
  \end{tablenotes}
  \label{tab:stats_by_nomination}
  \end{threeparttable}
\end{table}

% \begin{table}[H]
%   \centering
%   \begin{threeparttable}
%   \caption{Summary statistics for network members by referral status}
%   \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccccc}
%   \toprule
%   & \multicolumn{2}{c}{Verbal} & \multicolumn{2}{c}{Math} \\
%   \cmidrule{2-3} \cmidrule{4-5}
%   & Not Referred & Referred & Not Referred & Referred \\
%   \midrule
%   Reading z-score & 0.070 & 0.509 & 0.079 & 0.465 \\
%   & (0.003) & (0.039) & (0.003) & (0.040) \\
%   [1ex]
%   Math z-score & 0.079 & 0.452 & 0.087 & 0.590 \\
%   & (0.003) & (0.042) & (0.003) & (0.043) \\
%   [1ex]
%   GPA z-score & -0.066 & 0.705 & -0.069 & 0.711 \\
%   & (0.003) & (0.041) & (0.003) & (0.041) \\
%   [1ex]
%   Courses taken z-score & -0.153 & 2.690 & -0.184 & 2.488 \\
%   & (0.003) & (0.091) & (0.003) & (0.090) \\
%   [1ex]
%   Low-SES & 0.334 & 0.374 & 0.338 & 0.384 \\
%   & (0.001) & (0.019) & (0.001) & (0.019) \\
%   [1ex]
%   Med-SES & 0.515 & 0.513 & 0.513 & 0.507 \\
%   & (0.001) & (0.019) & (0.001) & (0.019) \\
%   [1ex]
%   High-SES & 0.151 & 0.113 & 0.149 & 0.109 \\
%   & (0.001) & (0.012) & (0.001) & (0.012) \\
%   % [1ex]
%   % Female & 0.524 & 0.562 & 0.520 & 0.428 \\
%   % & (0.001) & (0.019) & (0.001) & (0.019) \\
%   % [1ex]
%   % Age & 20.945 & 20.501 & 20.961 & 20.490 \\
%   % & (0.006) & (0.104) & (0.006) & (0.099) \\
%   \midrule
%   Observations & 128,174 & 673 & 127,481 & 669 \\
%   \bottomrule
%   \end{tabular*}
%   \begin{tablenotes}[flushleft]
%   \footnotesize
%   \item[] \textit{Note:} Standard errors in parentheses. GPA, test scores, and tie strength are standardized at the network level. For each referrer's network, we first calculated the mean and standard deviation of each measure. We then computed the average of these means and standard deviations across all referrers. Each individual's score was standardized using these network-level statistics. The standardization formula is $z = (x - \bar{x}_{network}) / \sigma_{network}$, where $\bar{x}_{network}$ and $\sigma_{network}$ are the average of network means and standard deviations, respectively. Low-SES, Med-SES, and High-SES are binary variables indicating the share of participants in estrato 1 and 2, 3 and 4, or 5 and 6, respectively. Tie strength measures the number of connections between individuals.
%   \end{tablenotes}
%   \label{tab:stats_by_nomination}
%   \end{threeparttable}
% \end{table}

@@

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Selection into the experiment}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc c@{}}
  \toprule
  & \multicolumn{1}{c}{\textbf{University}} & \multicolumn{1}{c}{\textbf{Sample}} & \multicolumn{1}{c}{\textit{\textbf{p}}} \\
  \midrule
  Reading score & 62.651 & 65.183 & 0.000 \\
  Math score & 63.973 & 67.477 & 0.000 \\
  GPA & 3.958 & 4.012 & 0.000 \\
  Low-SES & 0.343 & 0.410 & 0.000 \\
  Med-SES & 0.505 & 0.499 & 0.763 \\
  High-SES & 0.153 & 0.091 & 0.000 \\
  Female & 0.567 & 0.530 & 0.060 \\
  Age & 21.154 & 20.651 & 0.000 \\
  \midrule
  Observations & 4,417 & 734 &  \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} This table compares characteristics between the university and the experimental sample. $p$-values for binary outcomes (Low-SES, Med-SES, High-SES, Female) are from two-sample tests of proportions; for continuous variables, from two-sample $t$-tests with unequal variances. All reported \textit{p}-values are two-tailed.
  \end{tablenotes}
  \label{tab:selection}
  \end{threeparttable}
\end{table}

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Referral characteristics by academic area}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc c@{}}
  \toprule
   & \multicolumn{1}{c}{\textbf{Reading}} & \multicolumn{1}{c}{\textbf{Math}} & \multicolumn{1}{c}{\textit{\textbf{p}}} \\ 
  \midrule
  Reading score & 67.85 & 67.41 & 0.348 \\
  Math score & 70.04 & 71.36 & 0.029 \\
  Low-SES & 37\% & 38\% & 0.714 \\
  Med-SES & 51\% & 51\% & 0.829 \\
  High-SES & 11\% & 11\% & 0.824 \\
  \midrule
  N & 673 & 669 &  \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} This table compares characteristics of referred students by referrer's academic area (verbal vs. math). $p$-values are from two-sample t-tests with unequal variances. Math area participants refer peers with significantly higher math performance ($p = 0.029$), while no significant differences exist for reading scores or SES composition across areas.
  \end{tablenotes}
  \label{tab:referral_area}
  \end{threeparttable}
  \end{table}
